Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-02-10 text/paper2/ms.tex
added more intro text

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the major components of a galaxy are sensitive to the inner portion of its host halo's mass distribution. For example, the rotation curves of disk galaxies are particularly sensitive probes of the inner mass since matter in disks can be assumed to move on near circular orbits. This extra knowledge about the orbital distribution of the baryonic matter within a disk allows a more precise tracing of the potential than is feasible for objects such as elliptical galaxies where orbits are assumed to be random. [Potential is most likely smoothed out (energetic feedback?), closely integrable allowing beautiful dynamical ...]

@@

Within the Milky Way we have a unique vantage point that allows us to measure the three-dimensional structure of our own dark matter halo. One approach to ... [explain ``random'' methods: kinematic tracers]. Another approach is to exploit the non-random nature of the stellar populations in the halo. A consequence of hierarchical structure formation [...] and such structures have been observed [e.g., ...cite belokurov?]. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momenta --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might {\it assume} that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.][]{eyre}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g. along a stream), which would give us a direct measure of differences in a potential, $\Phi$.

Since the initial discovery of tidal debris from Sgr, myriad other streams and debris systems have been discovered in large photometric surveys probing Galactocentric distances from $\sim$15-100 kpc. This has motivated numerous efforts for developing methods to constrain properties of the Galactic potential with cold debris structures in the Milky Way halo.

- Broader comparison to other methods\\
\cite[][hereafter LM10]{lm10} ran N-body models of the disruption of the Sagittarius (Sgr) dwarf spheroidal galaxy and used these simulations to constrain the shape of the Milky Way potential out to $\sim$70 kpc [check number], finding that the halo ... (though, see ... for a discussion of unexplained features in the Sgr debris).
\indent\indent- 'generative methods' more generally (ie N-body, plus Varghese, plus Kuepper plus new possible action-angle models....\\
\indent\indent- statistical methods\\
\indent\indent- Sanders \& Binney 2013\\
- Then a short note on worries\\
\indent\indent- for generative models\\
\indent\indent- for any methods based on action/angle \\
\indent\indent- GD-1 was fit as if it highlighted an orbit.  That is known to be wrong.\\
\indent\indent- Densities are hard to match: need a background model, way to go from luminosity to density, needs assumptions about the mass function and structure of the satellite\\
\indent\indent- Triaxiality / radially-dependent shapes and substructure present problems for integral-based methods\\
- Then how rewinder gets round these????\\
\indent\indent- Reference Deg \& Widrow to say they illustrate the idea of the power of multiple streams to back up the bit where you say that's the long-term aim\\

>>>>

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the major components of a galaxy are sensitive to the inner portion of its host halo's mass distribution. For example, the rotation curves of disk galaxies are particularly sensitive probes of the inner mass since matter in disks can be assumed to move on near circular orbits. This extra knowledge about the orbital distribution of the baryonic matter within a disk allows a more precise tracing of the potential than is feasible for objects such as elliptical galaxies where orbits are assumed to be random. [Potential is most likely smoothed out (energetic feedback), can approximate as closely integrable allowing beautiful dynamical methods ...]

@@

Within the Milky Way we have a unique vantage point that allows us to measure the three-dimensional structure of our own dark matter halo. One approach to ... [explain ``random'' methods: kinematic tracers]. Another approach is to exploit the non-random nature of the stellar populations in the halo. A consequence of hierarchical structure formation [...] and such structures have been observed [e.g., ...cite belokurov?]. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might {\it assume} that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.][]{eyre}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g. along a stream), which would give us a direct measure of differences in a potential, $\Phi$. In practice, simple methods such as orbit fitting lead to systematic biases in inferred properties of the underlying potential because streams do not strictly delineate orbits [cite eyre binney work].

Since the initial discovery of tidal debris from Sgr, myriad other streams and debris systems have been discovered in large photometric surveys probing Galactocentric distances from $\sim$15-100 kpc. The known debris structures are $\gtrsim$10 kpc from the Sun, thus while overdensities in the halo are less contaminated by background, the constituent stars often have incomplete or poorly measured phase-space information. Perhaps the most rigorous method of modeling streams and the potential is to run full N-body simulations of satellite disruption and compare the density of observed stars to simulated stars. Such a model could naturally incorporate observational uncertainties, missing dimensions, and would, in principle, allow for studying any arbitrary time-dependent or non-integrable potential. However, such a model is presently computationally intractable [though see milkyway@home stuff...]. \cite[][hereafter LM10]{lm10} ran a grid of N-body models of the disruption of the Sagittarius (Sgr) dwarf spheroidal galaxy and matched their simulated streams to all Sgr data available at that time. By comparing the phase-space positions of their stream particles with the data, they were able to use these simulations to constrain the shape of the Milky Way potential out to $\sim$70 kpc [check number], finding that the halo ... (though, see ... for a discussion of unexplained features in the Sgr debris).

The difficulty and computational costs associated with proper N-body modeling has motivated numerous efforts for developing approximate methods for constraining properties of the Galactic potential with cold debris structures.

- Broader comparison to other methods\\
\indent\indent- 'generative methods' more generally (ie N-body, plus Varghese, plus Kuepper plus new possible action-angle models....\\
\indent\indent- statistical methods\\
\indent\indent- Sanders \& Binney 2013\\
- Then a short note on worries\\
\indent\indent- for generative models\\
\indent\indent- for any methods based on action/angle \\
\indent\indent- GD-1 was fit as if it highlighted an orbit.  That is known to be wrong.\\
\indent\indent- Densities are hard to match: need a background model, way to go from luminosity to density, needs assumptions about the mass function and structure of the satellite\\
\indent\indent- Triaxiality / radially-dependent shapes and substructure present problems for integral-based methods\\
- Then how rewinder gets round these????\\
\indent\indent- Reference Deg \& Widrow to say they illustrate the idea of the power of multiple streams to back up the bit where you say that's the long-term aim\\
